---
title: "Monarch occurrence .gif tutorial"
author: "Micah Freedman"
date: "7/30/2019"
output: html_document
---

##This is a quick tutorial on how to make an animated map of monarch butterfly occurrence data over the course of a year in North America. It uses data from iNaturalist (https://www.inaturalist.org/home) and is based on another tutorial from DataNovia (https://www.datanovia.com/en/blog/gganimate-how-to-create-plots-with-beautiful-animation-in-r/). 

###Begin by downloading observations from iNaturalist. Use this link: https://www.inaturalist.org/observations/export. For this demonstration, I specified only observations from North America and entered "Danaus plexippus" in the "Taxon" search field. Besides that, I left all of the default options. This dataset contains about 50,000 observations and took about 10 minutes to compile and download. Once it's ready, you will be directed to a link that has a zipped folder which contains a .csv file that has all of the data. The .csv file should have a name like "observations-59866.csv".

###Establish directory (put the downloaded .csv file here; all outputs will also go here)

```{r setup}
knitr::opts_knit$set(root.dir = '../monarch_map_gif/')
```

###Load necessary libraries; if you don't have these installed yet, you'll have to do that first using install.packages('name_of_package')

```{r}

library(ggplot2); library(plyr); library(rnaturalearth); library(gganimate); library(gapminder); library(gifski)

```

###Load the data and take a look. This could take a little while, depending on the size of the file.

```{r}

inat.monarchs <- read.csv(file = './observations-59866.csv')

head(inat.monarchs)

names(inat.monarchs)

```

###Filter out captive reared and non-research-grade observations.

```{r}

inat.monarchs <- inat.monarchs[inat.monarchs$captive_cultivated=='false',]

inat.monarchs <- inat.monarchs[inat.monarchs$quality_grade == 'research',]

```

###Pick which columns to keep

```{r}

columns.keep <- c('id','observed_on','quality_grade','place_guess','latitude','longitude','url')

inat.monarchs <- inat.monarchs[,columns.keep]

```

###Get date information into the proper format

```{r}

inat.monarchs$julian.date <- format(as.Date(inat.monarchs$observed_on), "%j") #create a new column for Julian date of observation

inat.monarchs$julian.date <- as.numeric(inat.monarchs$julian.date) #convert this to a continuous numeric variable

inat.monarchs$month <- format(as.Date(inat.monarchs$observed_on), "%m")

inat.monarchs$month <- month.name[as.numeric(inat.monarchs$month)] #convert 

inat.monarchs$month <- factor(inat.monarchs$month, levels = c('January','February','March','April','May','June','July','August','September','October','November','December')) #rearrange so months are in chronological order

```

###Specify observations to exclude. These are some observations that, despite being research grade, are not relevant, such as dead monarchs found in melted snow. I've almost certainly missed a few; this list can be updated.

```{r}
urls.exclude <- c('https://www.inaturalist.org/observations/9373451','https://www.inaturalist.org/observations/20483568','https://www.inaturalist.org/observations/9879408','https://www.inaturalist.org/observations/20320860','https://www.inaturalist.org/observations/21799902','https://www.inaturalist.org/observations/19494799','https://www.inaturalist.org/observations/11410855','https://www.inaturalist.org/observations/5740325','https://www.inaturalist.org/observations/7469853','https://www.inaturalist.org/observations/18866685','https://www.inaturalist.org/observations/18357363','https://www.inaturalist.org/observations/19379295')

inat.monarchs <- inat.monarchs[-(inat.monarchs$url %in% urls.exclude),] #exclude the above observations
```

###Plot observations onto a world map. Note: observations here are only from North America, since that is what I specified when I downloaded the data from iNaturalist.

```{r}

world <- ne_countries(scale= 'medium',returnclass = 'sf') #retrieve map layer (uses rnaturalearth package)

ggplot(data = world)+
  geom_sf()+
  theme_bw()+
  geom_point(data = inat.monarchs,  aes(x = longitude, y = latitude), size = 0.5, col = 'blue')+
  theme(legend.position = 'none')+
  theme(axis.title = element_blank())

```

###Okay, now it's time to make the actual .gif itself.

###First, we need to define a static plot that contains all of the data.

```{r}
(plot.static <- ggplot(data = world)+
  geom_sf()+
  theme_bw()+
  theme(axis.title = element_blank(), axis.text = element_blank())+
  xlim(c(-130,-60))+
  ylim(c(12,60))+
  theme(legend.position = 'none')+
  geom_point(data = inat.monarchs, 
             aes(x = longitude, y = latitude, col = julian.date), size = 1)+
  scale_color_viridis_c())
```

###Now use this map template to generate an animated .gif
####Arguments and what they do:
####transition_time() specifies what to use for indexing / what order to put plots in
####shadow_wake() will leave a "trail" or "imprint" behind after each point appears, with the time that it remains visible defined by the wake_length argument
####shadow_mark() will leave a permanent mark after a point appear.

```{r}
monarchs.anim <- plot.static + transition_time(julian.date) +
  shadow_wake(wake_length = 0.5, alpha = FALSE)+
  shadow_mark(alpha = 0.4, size = 0.6)
```

###Display the .gif!!!

```{r}
monarchs.anim
```

###Export the file in .gif format

```{r}
anim_save("../monarchs.gif", monarchs.anim, height = 8, width = 8, units = "in", res = 300)
```

###Make another plot, but this time plotting data according to their month of occurrence rather than Julian day

```{r}

(plot.static.month <- ggplot(data = world)+
    geom_sf()+
    theme_bw()+
    theme(axis.title = element_blank(), axis.text = element_blank())+
    xlim(c(-130,-60))+
    ylim(c(12,60))+
    geom_point(data = inat.monarchs, 
               aes(x = longitude, y = latitude, col = month), size = 1)+
    scale_color_viridis_d()) #same code as before, but with col = month and scale_color_viridis changed to discrete rather than continuous

monarchs.anim.month <- plot.static.month + transition_time(as.integer(month)) +
  shadow_wake(wake_length = 0.5, alpha = FALSE)+
  shadow_mark(alpha = 0.4, size = 0.6)+
  labs(title = 'Month: {frame_time}') #this line will add a title that shows Month: 1, Month: 2, etc.
  
monarchs.anim.month

```

